Spatial Reasoning and Statistical Graphics

### 2014-11-10 Paper/Draft1/TestingVisualAptitude.Rnw
Additional sentances explaining lineup task relationships. Some tweaks to figure sizes.

<<lineup-types,echo=F, fig.cap="Plot of scaled scores for different types of lineups\\label{fig:plottypesdensity}", fig.width=3, fig.height=6, out.width=".9\\linewidth", warning=F, message=F>>=
lineup.type.sum2$plot.type <- str_replace(lineup.type.sum2$plot.type, "rotated", "rot")
lineup.type.sum2$plot.type <- str_replace(as.character(lineup.type.sum2$plot.type), "error", "err")
lineup.type.sum2$plot.type <- factor(lineup.type.sum2$plot.type, 
                                     levels=c("violinplot", "jitter.boxplot", "bee.swarm.boxplot", "boxplot", 
                                              "scatterplot", "density", "histogram", 
                                              "qq.line", "qq.line.err", "qq.line.err.rot"), 
                                     labels=c("Violin Plot", "Boxplot + \nOverlaid Data", "Bee Swarm Boxplot", "Boxplot", 
                                              "Scatterplot", "Density Plot", "Overlaid Histogram", 
                                              "QQ plot + Line", "QQ plot +\nAcceptance Band", 
                                              "Rotated QQ plot +\nAcceptance Band")
                                     )
qplot(data=subset(lineup.type.sum2, !is.na(plot.type)& testtype=="lineup"), 
      x=score, y=..scaled.., colour=factor(testnum), geom="density", cut=4) + 
  facet_wrap(~plot.type, ncol=2) + 
  xlab("Proportion of Lineups Correct") + 
  ylab("Density") + 
  geom_vline(aes(xintercept=0), linetype=3) + 
  scale_colour_discrete("Lineup Test Number") + 
  scale_x_continuous(breaks=c(-.33, 0, .33, .67, 1)) +
  scale_y_continuous(breaks=c(0, .5, 1)) +
  theme_bw() + 
  theme(legend.position="bottom", legend.direction="horizontal", 
        axis.text.x=element_text(angle=30, hjust=1),
        plot.margin=unit(c(0,0,0,0),"mm"), 
        axis.ticks.y=element_blank(), axis.text.y=element_blank())
@
\newtext{
Figure \ref{fig:plottypesdensity} compares performance on each different type of plot. As two different lineup tasks utilized boxplots to test different qualities of the distribution of data (outliers vs. difference in medians), different tasks are shown as different colors, so that accuracy on tasks which are shown in blue can be compared to other blue density curves. 
}

\newtext{
Figure \ref{fig:scatterplottypes} shows the association between scaled score on each type of lineup and score on the visual reasoning tests. Sample size for each plot type is fairly small - between 5 and 10 plots per individual, so there is low power for systematic inference, but we can establish that the card rotation task is much more significantly associated with the QQ plots tasks compared to the other tasks. In addition, rotated qq plots seem to be much more asssociated with the paper folding task scores than other qq plot tasks; this may be because they require more visual manipulation than other qq plots?
}
<<lineup-types-scores,echo=F, fig.cap="Plot of scaled lineup scores by aptitude test scores.\\label{fig:scatterplottypes}", fig.width=6, fig.height=12, out.width="\\linewidth", fig.env="figure", warning=F, message=F>>=
lineup.type.sum3 <- dcast(lineup.type.sum2, id + testtype ~ plot.type, value.var="score", fun.aggregate = mean)
other.scores <- subset(lineup.type.sum3, testtype!="lineup")[,c(1,2,13)]
names(other.scores)[3] <- "testscore"
lineup.scores <- melt(subset(lineup.type.sum3, testtype=="lineup")[,c(1,3:12)], id.vars = 1)
names(lineup.scores)[2:3] <- c("lineup.type", "lineup.pct")
lineup.type.sum3 <- merge(other.scores, lineup.scores, all.x=T, all.y=T)
lineup.type.sum3$lineup.type <- factor(lineup.type.sum3$lineup.type, 
                                     levels=c("Violin Plot", "Boxplot + \nOverlaid Data", "Bee Swarm Boxplot", "Boxplot", 
                                              "Scatterplot", "Density Plot", "Overlaid Histogram", 
                                              "QQ plot + Line", "QQ plot +\nAcceptance Band", 
                                              "Rotated QQ plot +\nAcceptance Band"), 
                                     labels=c("Violin Plot", "Boxplot\n + Points", "Bee Swarm \nBoxplot", "Boxplot", 
                                              "Scatter\nPlot", "Density\nPlot", "Overlaid\nHistogram", 
                                              "QQ plot\n+ Line", "QQ plot\n+ Bands", 
                                              "Rotated QQ\nplot + Bands")
                                     )

@@

\section{Discussion and Conclusions}

>>>>

<<lineup-types,echo=F, fig.cap="Plot of scaled scores for different types of lineups\\label{fig:plottypesdensity}", fig.width=4, fig.height=8, out.width=".9\\linewidth", warning=F, message=F>>=
lineup.type.sum2$plot.type <- str_replace(lineup.type.sum2$plot.type, "rotated", "rot")
lineup.type.sum2$plot.type <- str_replace(as.character(lineup.type.sum2$plot.type), "error", "err")
lineup.type.sum2$plot.type <- factor(lineup.type.sum2$plot.type, 
                                     levels=c("violinplot", "jitter.boxplot", "bee.swarm.boxplot", "boxplot", 
                                              "scatterplot", "density", "histogram", 
                                              "qq.line", "qq.line.err", "qq.line.err.rot"), 
                                     labels=c("Violin Plot", "Boxplot + Points", "Bee Swarm Boxplot", "Boxplot", 
                                              "Scatterplot", "Density Plot", "Overlaid Histogram", 
                                              "QQ plot + Line", "QQ plot +\nAcceptance Band", 
                                              "Rotated QQ plot +\nAcceptance Band")
                                     )
qplot(data=subset(lineup.type.sum2, !is.na(plot.type)& testtype=="lineup"), 
      x=score, y=..scaled.., colour=factor(testnum), geom="density", cut=4) + 
  facet_wrap(~plot.type, ncol=2) + 
  xlab("Proportion of Lineups Correct") + 
  ylab("Density") + 
  geom_vline(aes(xintercept=0), linetype=3) + 
  scale_colour_discrete("Lineup Test Number") + 
  scale_x_continuous(breaks=c(-.33, 0, .33, .67, 1)) +
  scale_y_continuous(breaks=c(0, .5, 1)) +
  theme_bw() + 
  theme(legend.position="bottom", legend.direction="horizontal", 
        axis.text.x=element_text(angle=30, hjust=1),
        plot.margin=unit(c(0,0,0,0),"mm"), 
        axis.ticks.y=element_blank(), axis.text.y=element_blank())
@
\newtext{
Figure \ref{fig:plottypesdensity} compares performance on each different type of plot. As two different lineup tasks utilized boxplots to test different qualities of the distribution of data (outliers vs. difference in medians), different tasks are shown as different colors, so that accuracy on tasks which are shown in blue can be compared to other blue density curves. 
}

\newtext{
Figure \ref{fig:scatterplottypes} shows the association between scaled score on each type of lineup and score on the visual reasoning tests. Sample size for each plot type is fairly small - between 5 and 10 plots per individual, so there is low power for systematic inference, but we can establish that the card rotation task is much more significantly associated with the QQ plots tasks compared to the other tasks. In addition, rotated qq plots seem to be much more asssociated with the paper folding task scores than other qq plot tasks; this may be because they require more visual manipulation than other qq plots?
}
<<lineup-types-scores,echo=F, fig.cap="Plot of scaled lineup scores by aptitude test scores.\\label{fig:scatterplottypes}", fig.width=6, fig.height=12, out.width="\\linewidth", fig.env="figure", warning=F, message=F>>=
lineup.type.sum3 <- dcast(lineup.type.sum2, id + testtype ~ plot.type, value.var="score", fun.aggregate = mean)
other.scores <- subset(lineup.type.sum3, testtype!="lineup")[,c(1,2,13)]
names(other.scores)[3] <- "testscore"
lineup.scores <- melt(subset(lineup.type.sum3, testtype=="lineup")[,c(1,3:12)], id.vars = 1)
names(lineup.scores)[2:3] <- c("lineup.type", "lineup.pct")
lineup.type.sum3 <- merge(other.scores, lineup.scores, all.x=T, all.y=T)
lineup.type.sum3$lineup.type <- factor(lineup.type.sum3$lineup.type, 
                                     levels=c("Violin Plot", "Boxplot + Points", "Bee Swarm Boxplot", "Boxplot", 
                                              "Scatterplot", "Density Plot", "Overlaid Histogram", 
                                              "QQ plot + Line", "QQ plot +\nAcceptance Band", 
                                              "Rotated QQ plot +\nAcceptance Band"), 
                                     labels=c("Violin Plot", "Boxplot\n + Points", "Bee Swarm \nBoxplot", "Boxplot", 
                                              "Scatter\nPlot", "Density\nPlot", "Overlaid\nHistogram", 
                                              "QQ plot\n+ Line", "QQ plot\n+ Bands", 
                                              "Rotated QQ\nplot + Bands")
                                     )

@@

\newtext{For comparison, the correlation between general lineup score (non-subdivided) and the card rotation test score was \Sexpr{sprintf("%.3f", cor(ans.summary$lineup, ans.summary$card_rot))}, the correlation between general lineup score and the figure classification test was \Sexpr{sprintf("%.3f", cor(ans.summary$lineup, ans.summary$fig_class))}, and the correlation between lineup score and the paper folding test was \Sexpr{sprintf("%.3f", cor(ans.summary$lineup, ans.summary$folding))}.}
\newtext{While we can compare the correlation strength between tasks, it is clear that the correlation between the score on any single lineup type and a particular visual aptitude score is lower than the overall relationship that we attribute to visual ability. }
\newtext{Additional data is imperative to understand the reasoning required for specific types of plots - it is likely that the 5-10 trials per participant presented in each chart in figure \ref{fig:scatterplottypes} are simply not sufficient to uncover any specific relationship between reasoning ability and lineup task. 
}
\section{Discussion and Conclusions}
